### 2019-08-31 Manuscript/Manuscript.tex
Updates from Overleaf

Given that the region is a modern-day hot spot of fresh water and sediment discharge \citep{Milliman2013a} and chemical weathering and \COtwo consumption \citep{Hartmann2009a}, what effect has the emergence of I\&NG had on Earth's geochemical cycles and climate state? If this region was instead under water, would we have a glacial climate today? Based on paleogeographic reconstructions of arc-continent collisions through time, \citet{Macdonald2019a} proposed that arc-continent collisions in the tropics are responsible for increasing planetary weatherability, lowering \pCOtwo and causing the onset of glacial climate states. After the Miocene Climatic Optimum (MCO), a cooling trend began at $\sim$14~Ma associated with the growth of the Antarctic ice sheet, and accelerated over the past 4~m.y. culminating with the onset of northern hemisphere glaciation \citep{Shackleton1984a, Zachos2001a}. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for northern hemisphere glaciation being significantly lower at $\sim$280~ppm \citep{DeConto2008a}. The lower \pCOtwo threshold for northern hemisphere glaciation provides an explanation for southern hemisphere glaciation significantly preceding the development of ice sheets in both hemispheres. We seek to evaluate the hypothesis that the increase of subaerially-exposed land area in I\&NG has played a significant role in lowering \pCOtwo thereby driving this cooling and the development of northern hemisphere ice sheets.

@@

I\&NG are potent sinks of carbon because they contain large swaths of ophiolites that are exposed along E-W belts of tectonic topography in the warm wet tropics. In New Guinea, these paleogeographic features originated in the Middle Miocene with the collision between the Melanesian Arc and the distal margin of Australia \citep{vanUfford2005a, Cloos2005a, Baldwin2012a}, which resulted in the exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. The Middle Miocene (16-14~Ma) basal Markats Formation contains siliciclastic sediment that was transported from the south into the forearc basin associated with the New Guinea ophiolite belt \citep{Cloos2005a}. This precedes, by several million years, the beginning of widespread synorogenic sedimentation to the south and on the Australian continental basement at ca. 12~Ma \citep{vanUfford2005a}. Exhumation of the Central Range accelerated over the past 4~m.y. due to slab-breakoff and buoyant uplift under New Guinea \citep{Cloos2005a}. After arc-continent collision in New Guinea, a left-lateral strike-slip margin developed on the northern margin \citep{Baldwin2012a}. Over the past 4~m.y., this change in plate motion caused the westward extrusion and exhumation of the Cretaceous to Oligocene ophiolites of Obi, Halmahera, and northern New Guinea \citep{Baldwin2012a}. In eastern New Guinea, progressive jamming of the north-dipping subduction zone has caused major uplift over the past 4~m.y. \citep{vanUfford2005a}, which is well dated with a change in provenance from continental to volcanic detritus \citep{Abbott1994a} and thermochronology \citep{Hill1989a}.

@@

We thank Alec Brenner, Sam Lo Bianco, Mariana Lin, and Judy Pu for their contributions to the paleoshoreline reconstructions. Collaborative research between NLS-H and YG was initially supported by the France-Berkeley Fund. FAM and NLS-H thank NSF FRES grant \#20191023 for support.

>>>>

Given that the region is a modern-day hot spot of fresh water and sediment discharge \citep{Milliman2013a} and chemical weathering and \COtwo consumption \citep{Hartmann2009a}, what effect has the emergence of I\&NG had on Earth's geochemical cycles and climate state? If this region was instead under water, would we have a glacial climate today? Based on paleogeographic reconstructions of arc-continent collisions through time, \citet{Macdonald2019a} proposed that arc-continent collisions in the tropics are responsible for increasing planetary weatherability, lowering \pCOtwo and causing the onset of glacial climate states. After the Miocene Climatic Optimum (MCO), a cooling trend began at $\sim$14~Ma associated with the growth of the Antarctic ice sheet, and accelerated over the past 4~m.y. culminating with the onset of northern hemisphere glaciation \citep{Shackleton1984a, Zachos2001a}. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for northern hemisphere glaciation being significantly lower at $\sim$280~ppm \citep{DeConto2008a}. The lower \pCOtwo threshold for northern hemisphere glaciation provides an explanation for southern hemisphere glaciation significantly preceding the development of ice sheets in both hemispheres. To evaluate the hypothesis that the increase of subaerially-exposed land area in I\&NG has played a significant role in lowering \pCOtwo and the development of northern hemisphere ice sheets, here we use Neogene paleoshoreline reconstructions and couple a weathering model to a climate model to estimate changes \pCOtwo with changes in tropical paleogeography.

@@

I\&NG are potent sinks of carbon because they contain large swaths of ophiolites that are exposed along E-W belts of tectonic topography in the warm wet tropics. In New Guinea, these paleogeographic features originated in the Middle Miocene with the collision between the Melanesian Arc and the distal margin of Australia \citep{vanUfford2005a, Cloos2005a, Baldwin2012a}, which resulted in the exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. The Middle Miocene (16-14~Ma) basal Markats Formation contains siliciclastic sediment that was transported from the south into the forearc basin associated with the New Guinea ophiolite belt \citep{Cloos2005a}. This precedes, by several million years, the beginning of widespread synorogenic sedimentation to the south and on the Australian continental basement at ca. 12~Ma \citep{vanUfford2005a}. Arc-continent collision in New Guinea has progressed from west to east. Exhumation of the Central Range accelerated over the past 4~m.y. due to slab-breakoff and buoyant uplift \citep{Cloos2005a}. This was accompanied with a change to left-lateral lateral motion on the northern coast of New Guinea, which caused the exhumation of  ophiolites both along the coast and on the islands Obi and Halmahera \citep{Baldwin2012a}. In eastern New Guinea, progressive jamming of the north-dipping subduction zone has caused major uplift over the past 4~m.y. \citep{vanUfford2005a}, which is well dated with a change in provenance from continental to volcanic detritus \citep{Abbott1994a} and thermochronology \citep{Hill1989a}.

@@

We thank Alec Brenner, Sam Lo Bianco, Mariana Lin, and Judy Pu for their contributions to the paleoshoreline reconstructions. Collaborative research between NLS-H and YG was initially supported by the France-Berkeley Fund. FAM and NLS-H thank NSF FRES grant \#1926001 for support.
